---
title: "Case Study 1 How Does a Bike-Share Navigate Speedy Success?"
author: "Tineyi Kureva"
date: "2023-04-19"
output:
  word_document: default
  html_document: default
---
## ASK

In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.
Until now, Cyclistic’s marketing strategy relied on building general awareness and appealing to broad consumer segments. One approach that helped make these things possible was the flexibility of its pricing plans: single-ride passes, full-day passes, and annual memberships. **Customers who purchase single-ride or full-day passes are referred to as casual riders.** **Customers who purchase annual memberships are Cyclistic members**.
Cyclistic’s finance analysts have concluded that annual members are much more profitable than casual riders. Although the pricing flexibility helps Cyclistic attract more customers, Moreno believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all-new customers, Moreno believes there is a very good chance to convert casual riders into members. She notes that casual riders are already aware of the Cyclistic program and have chosen Cyclistic for their mobility needs.

### Business Task

Three questions have been asked by the business and they are:

* 1.How do annual members and casual riders use Cyclistic bikes differently?
* 2.Why would casual riders buy Cyclistic annual memberships?
* 3.How can Cyclistic use digital media to influence casual riders to become members?

As the Data Analyst I have tasked with providing answers on the first question: How do annual
members and casual riders use Cyclisric bikes differently?


## PREPARE

Data used for this analysis was obtained from <https://divvy-tripdata.s3.amazonaws.com/index.html>
This data has been made available by Motivate International Inc, under their license.
This is public data that can be used to explore how different customer types are using Cyclistic Bikes. No personal information including, names, credit card details  is available
The data obtained is in the form of 4 .csv files containing data for each Quarterly period of 2019..

### ROCCC Analysis

* Reliable - yes
* Original - yes
* Comprehensive - yes
* Current - yes
* Cited - yes

##PROCESS
### Wrangling
Libraries were loaded and the working directory set to access files
```{r}
#load libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)

getwd()

#set working directory
setwd("C:/Users/0002168344/Documents/destination_folder/Divvy_Exercises")

```
The 4 csv files were loaded for analysis
```{r upload Divvy datasets}
#upload Divvy datasets
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")

```

Column names were inspected and file q2_2019 had to be renamed to match the other csv files

```{r compare colum names}
#compare column names
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2019)
```
Columns were renamed in file q2_2019 to match other files
```{r rename columns}
#rename columns
q2_2019 <- rename(q2_2019
                  ,"trip_id" ="01 - Rental Details Rental ID" 
                  ,"start_time" = "01 - Rental Details Local Start Time"  
                  ,"end_time"= "01 - Rental Details Local End Time" 
                  , "bikeid" = "01 - Rental Details Bike ID"
                  , "tripduration" = "01 - Rental Details Duration In Seconds Uncapped"
                  ,"from_station_id" = "03 - Rental Start Station ID"
                  ,"from_station_name" = "03 - Rental Start Station Name" 
                  ,"to_station_id" = "02 - Rental End Station ID" 
                  ,"to_station_name" = "02 - Rental End Station Name" 
                  , "usertype" = "User Type" 
                  , "gender" = "Member Gender" 
                  , "birthyear" = "05 - Member Details Member Birthday Year")

```

Inspect data and look for incongruencies
```{r Inspect the dataframes and look for incongruencies}
str(q1_2019)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```
trip_id and bikeid column datatyp was converted to Chr to allow stacking of dataframes
```{r # Convert trip_id and bikeid to character so that they can stack correctly}
# Convert trip_id and bikeid to character so that they can stack correctly
q4_2019 <-  mutate(q4_2019, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid)) 
q3_2019 <-  mutate(q3_2019, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid)) 
q2_2019 <-  mutate(q2_2019, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid)) 
q1_2019 <-  mutate(q1_2019, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid)) 
```
All the 4 csv files are now merged into 1 data frame named all_trips
```{r }
# Stack individual quarter's data frames into one big data frame
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2019)
```
### Cleaning

Statistical summary of data
```{r}
summary(all_trips)
```
Checking for duplicate data in rows
```{r}
length(unique(all_trips$trip_id)) == nrow(all_trips)
```
Drop missing values
```{r}
all_trips <- all_trips %>% drop_na()
```
```{r}
#change subscriber customer to member and casual
all_trips <-  all_trips %>% 
  mutate(usertype = recode(usertype
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

```{r}
# Add columns that list the date, month, day, and year of each ride

all_trips$date <- as.Date(all_trips$start_time) #The default format is yyyy-mm-dd

all_trips$month <- format(as.Date(all_trips$date), "%B")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

## ANALYZE

```{r}
# Compare members and casual users
aggregate(all_trips$tripduration ~ all_trips$usertype, FUN = mean)
aggregate(all_trips$tripduration ~ all_trips$usertype, FUN = median)
aggregate(all_trips$tripduration ~ all_trips$usertype, FUN = max)
aggregate(all_trips$tripduration ~ all_trips$usertype, FUN = min)
```

```{r}
# See the average ride time by each day for members vs casual users
aggregate(all_trips$tripduration ~ all_trips$usertype + all_trips$day_of_week, FUN = mean)

```

```{r}
# Ordering the Days of the week.
all_trips$day_of_week <- ordered(all_trips$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

```{r}
#Average ride time by each day for members vs casual users
aggregate(all_trips$tripduration ~ all_trips$usertype + all_trips$day_of_week, FUN = mean)

```

```{r}
#Analyze ridership data by type and weekday
all_trips %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(usertype, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(tripduration)) %>% 		# calculates the average duration
  arrange(usertype, weekday)
```

## SHARE
### Visualizations and Key Findings
```{r}
#Number of rides by rider type
all_trips %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(tripduration)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge")
```

```{r}
#Average tripduration
all_trips %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(tripduration)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = usertype)) +
  geom_col(position = "dodge")
```

```{r}
#Average number of rides per month
all_trips %>% 
  group_by(month, usertype) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  filter(usertype == 'casual') %>%
  drop_na() %>%
  ggplot(aes(x = month, y = number_of_rides, fill = usertype)) + 
  geom_bar(position = 'dodge', stat = 'identity') + scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "Month", y = "Number of Rides", 
       fill = "usertype",
       title = "Average Number of Rides per month:Casual Rider")
```


```{r}
all_trips %>% 
  group_by(month, usertype) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  filter(usertype == 'member') %>%
  drop_na() %>%
  ggplot(aes(x = month, y = number_of_rides, fill = usertype)) + 
  geom_bar(position = 'dodge', stat = 'identity') + scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "Month", y = "Number of Rides", 
       fill = "usertype",
       title = "Average Number of Rides per month:Member Rider")
```


```{r}
#Create a dataframe for stations all_stations_v2, all_stations_member and all_stations_casual
all_stations <- bind_rows(data.frame("stations" = all_trips$from_station_name, 
                                     "usertype" = all_trips$usertype),
                          data.frame("stations" = all_trips$to_station_name,
                                     "usertype" = all_trips$usertype))
all_stations_v2 <- all_stations[!(all_stations$stations == "" | is.na(all_stations$stations)),]
all_stations_member <- all_stations_v2[all_stations_v2$usertype == 'member',]
all_stations_casual <- all_stations_v2[all_stations_v2$usertype == 'casual',]
```


```{r}
#Top 10 stations
top_10_station <- all_stations_v2 %>% 
  group_by(stations) %>% 
  summarise(station_count = n()) %>% 
  arrange(desc(station_count)) %>% 
  slice(1:10)
top_10_station


```


```{r}
#Top 10 stations for member riders
top_10_station_member <- all_stations_member %>% 
  group_by(stations) %>% 
  summarise(station_count = n()) %>% 
  arrange(desc(station_count)) %>% 
  head(n=10)
top_10_station_member

```


```{r}
#Top 10 stations for casual riders
top_10_station_casual <- all_stations_casual %>% 
  group_by(stations) %>% 
  summarise(station_count = n()) %>% 
  arrange(desc(station_count)) %>% 
  head(n=10)
top_10_station_casual
```

```{r}
#Average rides by hour
str(all_trips)
all_trips$start_time <- as.POSIXct(all_trips$start_time, "%Y-%m-%d %H:%M:%S")      
str(all_trips)
```

```{r}
#Average riders per hour for casual riders
all_trips %>%
  filter(usertype == 'casual') %>%
  group_by(hour_of_day = hour(round_date(start_time, 'hour'))) %>% 
  group_by(hour_of_day, usertype) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  arrange(-number_of_rides) %>% 
  ggplot(aes(x = hour_of_day, y = number_of_rides, fill = usertype)) +
  geom_bar(position = 'dodge', stat = 'identity') + scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23)) +
  labs(x = "Time of the Day (h)", y = "Number of Rides", 
       fill = "Casual",
       title = "Average Number of Rides per Hour: Casual Riders")

```

```{r}
#average riders per hour for casual riders
all_trips %>%
  filter(usertype == 'member') %>%
  group_by(hour_of_day = hour(round_date(start_time, 'hour'))) %>% 
  group_by(hour_of_day, usertype) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  arrange(-number_of_rides) %>% 
  ggplot(aes(x = hour_of_day, y = number_of_rides, fill = usertype)) +
  geom_bar(position = 'dodge', stat = 'identity') + scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23)) +
  labs(x = "Time of the Day (h)", y = "Number of Rides", 
       fill = "Member",
       title = "Average Number of Rides per Hour: Member Riders")
```


```{r}
#Bike Usage between Members and Casual Riders by Day of Week across the Year
options(repr.plot.width = 35, repr.plot.height = 10)
all_trips %>% 
  group_by(month, day_of_week, usertype) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  drop_na() %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::comma) +
  facet_grid(usertype~month) +
  labs(x = "Day of Week", y = "Number of Rides", fill = "Member/Casual",
       title = "Bike Usage between Members and Casual Riders per Day of Week for 2019", fill = 'Member/Casual') +
  theme(axis.text.x = element_text(angle = 90))
```

### Key Finding

* It is evident that there are on average there are more Casual Riders compared to Member riders and that Casual Riders trip duration is longer than Member Riders
* Casual riders per hour outnumber Member riders. From the bar graph; Member riders seem to  have 2 peaks name from 6am-10am and 4pm-7pm which may suggest that Member riders use the service to take them to and from work
* The top 3 stations for Casual Riders are Streeter Dr & Grand Ave,Lake Shore Dr & Monroe St   and Lake Shore Dr & North Blvd. 
  Coursehero website says "Streeter Dr. & Grand Ave. is near a very popular pier in Chicago   and a Museum, I would conclude people use the bikes for fun and to get to and between     points of interest. Lake Shore Dr. & Monroe St. and Lake Shore Dr. & North Blvd. are near Grant Park which is a great place to go bike riding,"<https://www.coursehero.com/file/p335u4tb/Streeter-Dr-Grand-Ave-is-near-a-very-popular-pier-in-Chicago-and-a-Museum-I/>. This explains the popularity of Casual riders and the use being for leisure/tourist activities.
  
## ACT

### Top Three Recomendations
 * A digital marketing strategy to attract Casual Riders in the 3 Most popular stations and make them aware that they can convert to membership status and use the bikes for other purposes like work and grocery shopping.
 * Offer discounts to member riders who refer casual riders
 * Offer weekend membership